---
title: "tRaining Day 1"
author: "ZS"
date: "June 8, 2018"
output: word_document
editor_options: 
  chunk_output_type: console
---
```{r setup}
rm(list=ls())
```


### Back to Basics
To run the example of a fucntion from a given package, use example(function_name)

 - To run an entire script, shift+ctrl+enter
 - Running everything before your line, ctrl+alt+P (in RMarkdown, it will run the previous chunk)

```{r Basics}
# Starting Code----
example(mean)
rm(x)
ls()

# Getting Help----
# Get inforamtion about the function
mean

# Say you wan thelp with an arithmetic operator, cannot use ?function to get help
help("+") # Will pull up the page for Arithmetic functions

# Environment----
# use ls() to see what's in your environment
ls()
rm(list=ls()) # clears everything from your environment, should be used often when running a script so you're not working with old objects
ls()

#' Will allow you to carry the first line - Markdown equivalent?
```

### Data Structures

Think about *mode* and *class*

 - *Vector*: could be a list of things (1 dimension), all of which belong to *one* mode (e.g. true/false, letters, numbers)
 - *Factor*: 1 dimensional holds categorical values (e.g. colors, treatments)
 - *Matrix*: 2 dimensional, but all things are of 1 mode (all numbers, letters, etc)
 - *Array*: can be multidimensional, but all things are all still same mode
 - *List*: 1 dimension, but each thing in the list can be different modes
 - *Data.Frame*: 2  dimensional, each column is an element, but they can be different modes
    - BUT must have equal row/column lengths to have the df
    - Think of it as a list, transposed (where list lenghts are the same)
    
#### Vectors

Think of things in terms of "elements", each thing in an object

```{r Creating Vectors}
# Creating a vector
x <- 66:12
str(x) # Returns int [1:55] since there are 55 elements
class(x) # Double precision factor
mode(x) # Returns "numeric"

# Using the "c" function
x <- c(4, 132, 48) # Creating a number vector
X <- c( "word", "lizard", "beaker") # Creating a character vector,  length of 3
x <- c("a sentence is a character", "this is a single, element") # length of 2

```

#### Indexing

Three ways to index data: numeric, character, or logical

```{r Indexing Examples}
# Numeric Indexing----
# What goes between the [] is always a vector

# Create an object, x, with integers from 21 to 30
x <- 21:30

x[5] # Returns the 5th element

# Can use vector to index within an object
y <- 5
x[y] # Will also return 25

# R will index in the order you ask for information
x[c(2, 4, 7, 1)]

# If you don't know how many elements are in your vector
length(x) # Tells you the length of your vector
x[length(x)] # Allows you to index the last value without knowing how long the vector is

# Can assign the output of a function to an object
l.x <- length(x)
x[l.x] # show the last element of the vector

# Show first and last elements
i <- c(1, l.x)
x[i]

# Assign 1st and last elements to y
y <- x[i]

x <- 55:100
x # Does the same as print(x)

# Remove 18th element (72)
x[-18] # returns the vector without that value
x[c(-1, -18, -35)] # returns the vector without those three values
x[-c(1, 18, 35)] # does the same thing as line above
x[-(1:5)] # returns vector elements wihtout first 5 elements

# These two functions do the same thing
x <- x[-(1:5)] # Creates x by deleting the first 5
x <- x[6:length(x)] # Creates x by only taking 6th element to end

# Assign to an element
x[15] <- 15 # Change the 15th element to the integer 15
x[1:5] <- 5 # Change elements 1-5 to the number 5

# Character Indexing----
# Indexing by names only works if your vector has names (column names or as below)
z <- c("a", "b", "c", "d")
str(z)
letters # An object available by R

# Make a vector
x <- 1:10
names(x) # Now, would print "NULL" since there are no names assigned yet

# Assign names
names(x) <- letters[1:10]
str(x)

# Can use the names of values to get the values
x["f"] # Where "f" is a vector with length 1
x["i"] <- 100 # Looks for element named "i" and changes the element value to 100

z <- c("e", "i", "a") # Create a character vector
x[z] # Returns the values from the character vector

# Can only use character vectors to return those elements you want, not take them away
x[-"a"] # Will not work

names(x)[c(3:5)] <- c("sarah", "mike", "john") # Change names for element x
names(x)

# Logical Indexing----
# Result of a logical opration is a vector

# Can ask logical across entire vector
x > 3

# Index using a logical
x[x > 3] # Returns elements that are TRUE
x[x > 3 & x < 250] # same as 250 > x > 3
                    # Logical vector made up of two logical vectors, both of which must be true to return
x[x > 3 | x < 250] # same as return any thing x >3 OR x < 250

# Any and All in a logical vector
any(x) # Are any true for the condition
all(x) # Are all true
x[x] # Returns values that are true
```

#### Factors

```{r}
# Factors take characters, find uniue values, represent as numbers
# Easier to store as numbers, and rep the factor levels as numbers, using those as a sort of "lookup table"
# Levels are sorted Alphabetically and assigned numbers in order

x <- factor(c("yellow", "green", "blue", "red", "Blue", "yellow"))
x
str(x) # Factor with 5 levels
mode(x) # Numeric
class(x) # Factor

is.numeric(x) # Returns false, not a number, but is stored as one for reference
is.character(x) # Returns false

as.numeric(x) # To return numbers that it's assigned level factors
as.character(x)

x <- factor(c("yellow", "green", "blue", "red", "Blue", "yellow"),
            levels = c("yellow", "Blue", "red", "blue")) # you control the order of assignment to levels
```

#### Matrices

*Always* 2 dimensional things with rows and columns

```{r}
# Initialize with vector----
x <- letters[1:24]
mat <- matrix(x, nrow = 4)
mat # We've created a character matrix
length(mat) # How many elements are in mat
nrow(mat)
ncol(mat)

# Indexing in Matrix----
# mat[row, column]
mat[2, 1:5] # Drops the dimension and returns a character vector
mat[2, 1:5, drop = FALSE] # Maintains the dimensions and returns a one-row matrix

# Want first two clumns, but not 3 to 5, can mix positive and negative between slots, but not within
mat[1:2, -c(3, 5)]

# Make column 5 all x's
mat[,3] <- "x"
mat

# Assign row and column names
rownames(mat) <- LETTERS[26:23]
colnames(mat) <- LETTERS[1:6]
mat
str(mat)

# Now you can refer to rows and columns by names, rather than numbers
mat["Z", 1:5] # Can mix between using names

mat[c("Z","Y"),c("E","F")]

# Logical Indexing----
x <- 1:24
n.mat <- matrix(x, nrow = 6)
n.mat
# Choose all values that meet a condition, want all rows where column
i <- n.mat[,1] < 4
n.mat[i,] # Only returns rows where values in  column are less than 4

# Want columns where value across rows is greater than 15
c3gt15 <- 
  
# Index Rows
# rows wehre column two is less than 11 and greater than 9
  
cond <- n.mat[,2] > 9 & n.mat[,2] < 11 # Could use >, >=, or | too 
n.mat[cond,]

# Indexing Columns : can be done, but a lot less common
# Putting condition in [row,] you're trying to select rows
# Putting condition in [,column] you're trying to select columns

x <- letters[1:24]
mat <- matrix(x, nrow = 4)
mat

# Can use conditions from indexing on one matrix to index another, so long as 
# the lengths are the same length
mat
z <- c(15, 205, 3, 6, 1, -5)
z
mat[, z < 10] # Can only be accomplished bc the condition in [,column] is logical
z < 10 # This logical vector means that the last 4

# Adding rows or columns (default is to the end)
cbind(mat, c("Z", "Z", "A","A"))
rbind(n.mat, 101:104)

# Insert a row somewhere that's not at the end
rbind(n.mat[1:3,], c(5, 5, 5, 5), n.mat[4:6,])

# Matrices as square things where everything is the same (all #s)
```

#### Lists

Keeping things that are in different modes together in one place
Lists are essentially vectors, where elements are of different modes

```{r Lists}

x <- list(1, letters[5:10], matrix(1:24, nrow = 4))
x

# Removing something from a list, use [[]] to get the value or $ if it list elements have names
# [] will return the element of the list
y <- x[[3]]
y

x[[3]][3:4,] # First part is the matrix, 2nd part is indexing that matrix

# Adding Names to elements in a list
names(x) <- c("first", "second", "third")
x

x$second

# Can create list and assign names simultaneously
z <- list(num = 1, lets = letters[5:10], mat = matrix(1:24, nrow = 4))

# Can edit things in the lsit
z$mat[3, 1] <- 100
z$mat

# Can delete things
z$lets <- NULL
z

# Can add things
z$new.element <- TRUE
z

# List of lists
z$new.list <- list(chuck = "chuck", sally = matrix(1:10, nrow = 2))

example(lm) # Navigating through a list to get to what you want
names(lm.D9)
lm.D9$coefficients
lm.D9$coefficients["(Intercept)"] # Look in lm.D9 for "intercept" value within coefficients
```

#### Dataframes

```{r DFs}
rm(list = ls())

# Create four vectors
ids <- c(23, 345, 234, 565)
loc <- c("north", "north", "south", "west")
len <- c(9.9, 3.4, 4.6, 8.9)
wght <- c(234, 345, 122, 233)

df <- data.frame(ids = ids, loc = loc, len = len, wght = wght)
str(df)
df
nrow(df)
df[1] # Single brackets return a data frame
df[[1]] # Double brackest return elements of column 1
df$ids # Returns the same information as above

# Exxtract first column - done 4 ways
df[[1]]
df[, 1]
df$ids
df[["ids"]]

# Call out row 1 - two ways
df[1,] # calls first row
df["1",] # calls row called "1"

# Change row names
rownames(df) <- c("hal", "sue", "max", "jim")
df
df["hal",]

# **YOu can mix and match**
df[c(2, 4), c("len", "wght")]

# Rows where length is >3
cond <- df$len > 3 # This is your logical vector
df[cond,] # Returns the rows where the answer to the logical vector is true

# 3 > weight >100
df[df$len >3 & df$wght >100, ]


```

```{r R's Datasets}
?datasets
library(help = "datasets") # ALL THE DATASETS IN R, listed

head(iris) # Shows variable names, and first 6 objets
tail(iris) # Shows last 6 objects
summary(iris) # Summary data

iris[iris$Sepal.Length < 4.8, ]
iris$Sepal.Length < 4.8 # The logical vector that's being run

# Extract just the "setosa" species
iris.s <- iris[iris$Species == "setosa", ]
head(iris.s)

# Extract based on product of two columns, petal area = length x width
petal.area <- iris$Petal.Length * iris$Petal.Width
petal.area # not physically connected to iris df, has same number of values as two columns that went into it
iris[petal.area > 14, ]

# SUBSET function takes the df and allows you to write a condition
subset(iris, Species == "virginica" | Species == "setosa")
subset(iris, Species == "virginica" | Species == "setosa", c("Sepal.Length", "Species")) # using the 3rd argument to read back certain columns

```

#### Coercion

```{r}
chr <- as.character(1:5) # a vector of characters
num <- as.numeric(chr) + 10

chr <- c(1:5, "hello") # makes a character vector
as.numeric(chr) # Will print results, but present warming that there were NAs introduced by coercion

as.logical(c("TRUE", "true", "T", "True", "hello")) # as long as the whole word is there, it will convert, but not with "hello"

# R will convert numbers to logicals
as.logical(-5:5) # Any number that is NOT 0, whether positive or negative, is TRUE
as.numeric(as.logical(-5:5)) # Converts trues to 1, and falses to 0

str(iris)
# How many values of Sepal.Length are > 5
slgt5 <- iris[iris$Sepal.Length >5,]
str(slgt5)
nrow(slgt5) # gives you the count of rows the meet the conditions

sum(iris$Sepal.Length >5) # Where the (thing inside the paren) is a logical vector, and TRUEs are converted into 1, then sum() takes the total of those 1s

mean(iris$Sepal.Length >5) # Proportion of items that meet the condition
```

#### Missing Data

Cannot test for NA using the "=" argument

```{r}
z <- c(1, NA, 3, NA, 5, 7, NA)
is.na(z)
z[!is.na(z)] # Shows values that are not NA
!is.na(z) # Shows logical results for your vector

na.omit(z)

z1 <- z[!is.na(z)]

# Attributes are like metadata that you can attacth to objects that will stay with them even if you modify the object itself

iris.na <- iris
iris.na[1, "Sepal.Length"] <- NA
iris.na[8, "Petal.Length"] <- NA

# Want all rows where there is no missing data
# Complete cases----
# looks across row, returns TRUE if all columns contain values (no NAs)
iris.comp <- iris.na[complete.cases(iris.na), ]
head(iris.comp) # Deleted rows that contained one or more NAs

# Complete cases works only on the data frame you give it, so be sure to give it the data frame you care about, esp if you need to subset columns/rows first
iris.comp <- iris.na[complete.cases(iris.na[, c(1,2,4)]), ]
head(iris.comp)
```

#### Directories/Files

```{r}
getwd() # Tells you where your working directory is
dir()
dir("..") # Takes you one folder level up from your current directory
set("..") # Sets it one step up

# Writing Files---

# Functions for writing an object and reading it are paired and similar

# WOrkspace objects----------------- 
# (save and load)
save.image() # Makes an R data file that saves your workspace (data, objects) into your working directory
rm(list = ls()) # clear the workspace
load(".Rdata")

# To be specific about what you're saving
save(chr, iris.na, z, file = "stuff.rdata")
rm(list = ls())
load("stuff.rdata")
stuff <- load("stuff.rdata") 
stuff # a way to check what stuff it saved to the .rdata image

# CSV files or TabDelimited Values------------------ 
# (read.table and write.table/write.csv)
str(iris)
write.csv(iris, file = "iris.csv") # creates a comma-delimited file, with a rowname column by default
write.csv(iris, file = "iris.csv", row.names = F) # writes without column of row names

iris.2 <- read.csv("iris.csv") # must assign to an object to be able to manipulate it as a dataframe

# R will try to guess that things that are character values as factors, even if they aren't
iris.3 <- read.csv("iris.csv", stringsAsFactors = F)
str(iris.3)

# R codes/commands--------------
# dump and source
# dump creates a text file that contains the commands to create the objects

X <- 1:10
dump("X", file = "X.r")
rm(X)
source("x.r") # Goes to the command in X.r and run the command to create the object

```

1. Compute the following values:
• 27 times 38 minus 17
• natural logarithm of (56 divided by 4)
• square root of (4 times 13)
• 6 squared divided by 2
```{r 1}
27 * 38 - 17

log(56/4)

sqrt(4 * 13)

(6^2)/2
```

2. Create two vectors:

```{r 2}
# • 160, decreasing sequential integers by 10, 20, 15, 10, 5
x <- c(160, seq(150, 20, -10), 15, 10, 5)
x # building a vector by inserting functions that create other vectors

# could also do:
x <- c(160, seq(150, 20, -10), seq(15, 5, -5))
x

# • 57, 58, 59, sequential integers by 2, 85, 86
y <- c(57, 58, 59, seq(60, 84, 2), 85, 86)
y
# could also do:
y <- c(57:59, seq(60, 84, 2), 85:86)
y

# (If you can’t create these two vectors, then create one vector from 20 to 35 
# and another from 205
# to 190)


```

3. Multiply the two vectors above and assign to a new vector.
```{r}
z <- x * y
```

4. What are the 3rd, 10th, and 13th elements of the vector from 3?
```{r}
z[c(3, 10, 13)]
```

5. How many elements of the vector in 3 are greater than 6000?
```{r}
gt6k <- z[z>6000] # logical vetor that runs on all 18 elements from z
length(gt6k)

# Or, can use coersion
as.numeric(z > 6000)
sum(z > 6000)

# To print those values that are true for the logical z > 6000
as.numeric(gt6k)
```

6. What is the mean, median, and sum of the vector in 3?
```{r}
mean(z)
median(z)
sum(z)

# Can place them all in a summary vector, and give each object names as info
smry <- c(
  mean = mean(z),
  median = median(z),
  sum = sum(z)
)
smry
```

7. Change the 4th and 8th elements of the vector in 3 to 4 and 8.
```{r}
z[4] <- 4 # 4th element is 4
z[8] <- 8

# can accomplish in one step
z[c(4, 8)] <- c(4, 8)
z

# can also run it from a vector
new.vals <- c(4,8)
z[new.vals] <- new.vals
```

8. Using the USArrests data.frame, extract a vector of the percent of the population living in urban areas
(UrbanPop). Create names for this vector from the states in the row names of the data.frame.
```{r}
USArrests <- datasets::USArrests
urbanPop <- USArrests$UrbanPop # WILL NOT WORK IN MATRIX
urbanPop <- USArrests[,3]
urbanPop <- USArrests[, "UrbanPop"]
urbanPop <- USArrests[, c(F, F, T, F)]
urbanPop <- USArrests[["UrbanPop"]] # WILL NOT WORK IN MATRIX
urbanPop <- USArrests[[3]]

# Know the object you're working with, double brackets and $column name will not work with extracting
# use str(df) to determine if it's a df or a matrix

# Create names for this vector from the states in the row names of the df
row.names(USArrests)
colnames(USArrests)
names(urbanPop) <- rownames(USArrests) # Give the vector names and use that to extract the information for number 9
head(urbanPop)
```

9. What is the percentage of the population in urban areas for Kansas, California, and Kentucky?
```{r}
urbanPop[row.names(USArrests) == "Kansas" | row.names(USArrests) == "California" | row.names(USArrests) == "Kentucky"]
```

10. What is the average urban population percentage in New England (Maine, Vermont, New Hampshire,
Massachusetts, Rhode Island, and Connecticut)?
```{r}
names(urbanPop) <- rownames(USArrests)
neweng <- urbanPop[c("Maine", "Vermont", "New Hampshire", "Massachusetts", "Rhode Island", "Connecticut")]
mean(neweng)
```

11. What is the average murder/assault ratio in the entire USArrests data.frame?
```{r}
m.a.ratio <- USArrests$Murder / USArrests$Assault
mean(m.a.ratio)
```

12. Create a copy of the USArrests data.frame and add a column for the murder/assault ratio.
```{r}
USArrests$m.a.ratio <- USArrests$Murder / USArrests$Assault
mean(USArrests$m.a.ratio)
```

13. Write a .csv file of this modified data.frame.
```{r}
write.csv(USArrests, "USArrests.copy.csv") # assigns row names (states) to first column

# If you don't want row names
write.csv(USArrests, "USArrests.copy.1.csv", row.names = F)

# IF you want the row names and want to name that first column (default is no name)
USArrests$states <- row.names(USArrests)
write.csv(USArrests, "USArrests.copy.2.csv", row.names = F)
```

14. Read the .csv file from 13 to a new data.frame.
```{r}
csv.df <- read.csv("USArrests.copy.2.csv")
```

15. Extract rows from the data.frame read in 14 that have a murder/assault ratio less than the mean.
```{r}
mean.ma <- mean(csv.df$m.a.ratio)
lt.mean <- csv.df$m.a.ratio < mean.ma # Creates a logical vector of true false
lt.mean.df <- csv.df[lt.mean, ] #extracts the rows that are TRUE
lt.mean.df

# or, DO IT ALL IN ONE LINE!
lt.mean.df <- csv.df[csv.df$m.a.ratio < mean(csv.df$m.a.ratio), ]
lt.mean.df
```

